Endogenous Macrodynamics in Algorithmic Recourse

### 2022-11-03 paper/sections/limitations.rmd
incorporated some revsions

The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion. In practice, models are in fact updated on a regular basis [@upadhyay2021robust]. We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time $t=0$. That being said, our experimental design is a vast over-simplification of potential real-world scenarios. In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in @upadhyay2021robust. We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class. Relating this back to the consumer credit example, we assume that the would-be borrowers are always willing and able to implement recourse and the bank is always willing to provide credit as would-be borrowers move across the decision boundary. In practice it is doubtful that agents behave according to such simple rules. Nonetheless, we think that our simple framework offers a starting point for future work on recourse dynamics (both endogenous and exogenous dynamics).

>>>>

The experimental setup proposed here is designed to mimic a real-world recourse process in a simple fashion. In practice, models are in fact updated on a regular basis [@upadhyay2021robust]. We also find it plausible to assume that the implementation of recourse happens periodically for different individuals, rather that all at once at time $t=0$. That being said, our experimental design is a vast over-simplification of potential real-world scenarios. In practice, any endogenous shifts that may occur can be expected to be entangled with exogenous shifts of the nature investigated in Upadhyay et al. @upadhyay2021robust. We also make implicit assumptions about the utility functions of the involved agents that may well be too simple: individuals seeking recourse are assumed to always implement the proposed Counterfactual Explanations; conversely, the agent in charge of the model $M$ is assumed to always treat individuals that have implemented valid recourse as if they were truly now in the target class.

## Causal Modelling

In this work we have focused on popular counterfactual generators that do not incorporate any causal knowledge. The generated perturbations therefore may involve changes to variables that affect the outcome predicted by the black-box model, but not the true outcome. In the context of Example \@ref(exm:student), the generators may, for example, propose that students improve their results on some aptitude test, that is recognized as a signal by the university's hiring committee. Changes to those test scores can be expected to affect the underlying machine learning model even if there is no causal link between test outcomes and students' chances of graduating. The implementation of such changes is typically described as **gaming** @miller2020strategic. Preventing such gaming may help to avoid the macrodynamics we have pointed to in this work. Future work would therefore likely benefit from including recent approaches to AR that incorporate causal knowledge such as Karimi et al. @karimi2021algorithmic.

### 2022-11-03 paper/sections/methodology.rmd
incorporated some revsions

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$. To quantify predictive uncertainty, Antoran et al. [@antoran2020getting] rely on entropy estimates for probabilistic models. The greedy approach proposed by Schut et al. @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. [@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$, we also implicitly minimize predictive entropy (as in CLUE). In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by @dombrowski2021diffeomorphic relies on normalizing flows.].

>>>>

CLUE distinguishes itself from REVISE and other counterfactual generators in that it aims to minimize the predictive uncertainty of the model in question, $M$. To quantify predictive uncertainty, Antoran et al. [@antoran2020getting] rely on entropy estimates for probabilistic models. The greedy approach proposed by Schut et al. @schut2021generating, which we shall refer to as **Greedy**, also works with the subclass of models $\tilde{\mathcal{M}}\subset\mathcal{M}$ that can produce predictive uncertainty estimates. The authors show that in this setting the cost function $\text{cost}(\cdot)$ in Equation \@ref(eq:general) is redundant and meaningful counterfactuals can be generated in a fast and efficient manner through a modified Jacobian-based Saliency Map Attack (JSMA). Schut et al. [@schut2021generating] also show that by maximizing the predicted probability of $x^\prime$ being assigned to target class $y^*$, we also implicitly minimize predictive entropy (as in CLUE). In that sense, CLUE can be seen as equivalent to REVISE in the Bayesian context and we shall therefore refer to both approaches collectively as **Latent Space** generators^[In fact, there are a number of other recently proposed approaches to counterfactual search that also broadly fall in this same category. They largely differ with respect to the chosen generative model: for example, the generator proposed by Dombrowski et al. @dombrowski2021diffeomorphic relies on normalizing flows.].

### 2022-11-03 paper/sections/methodology_2.rmd
incorporated some revsions

where $X=\{x_1,...,x_m\}$, $\tilde{X}=\{\tilde{x}_1,...,\tilde{x}_n\}$ represent independent and identically distributed samples drawn from probability distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$ respectively @gretton2012kernel. MMD is a measure of the distance between the kernel mean embeddings of $\mathcal{X}$ and $\mathcal{\tilde{X}}$ in a Reproducing Kernel Hilbert Space, $\mathcal{H}$ [@berlinet2011reproducing]. An important consideration is the choice of the kernel function $k(\cdot,\cdot)$. In our implementation we make use of a Gaussian kernel with a constant length-scale parameter of $0.5$. As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD(X,\tilde{X})=0$ if and only if $X=\tilde{X}$.

The evaluation metric in Equation \@ref(eq:mmd) is computed after every round $t=1,...,T$ of the experiment. To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution, we follow @arcones1992bootstrap. To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$. Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples. Then under the null hypothesis, we should have that $MMD(X^\prime,\tilde{X}^\prime)$ be approximately equal to $MMD(X,\tilde{X})$. The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.

>>>>

where $X=\{x_1,...,x_m\}$, $\tilde{X}=\{\tilde{x}_1,...,\tilde{x}_n\}$ represent independent and identically distributed samples drawn from probability distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$ respectively @gretton2012kernel. MMD is a measure of the distance between the kernel mean embeddings of $\mathcal{X}$ and $\mathcal{\tilde{X}}$ in a Reproducing Kernel Hilbert Space, $\mathcal{H}$ [@berlinet2011reproducing]. An important consideration is the choice of the kernel function $k(\cdot,\cdot)$. In our implementation we make use of a Gaussian kernel with a constant length-scale parameter of $0.5$. As the Gaussian kernel captures all moments of distributions $\mathcal{X}$ and $\mathcal{\tilde{X}}$, we have that $MMD(X,\tilde{X})=0$ if and only if $X=\tilde{X}$. Conversely, larger values $MMD(X,\tilde{X})>0$ indicate that it is more likely that $\mathcal{X}$ and $\mathcal{\tilde{X}}$ are different distributions. In our context, large values therefore indicate that a domain shift indeed seems to have occurred.

To assess the statistical significance of the observed shifts under the null hypothesis that samples $X$ and $\tilde{X}$ were drawn from the same probability distribution, we follow @arcones1992bootstrap. To that end, we combine the two samples and generate a large number of permutations of $X + \tilde{X}$. Then, we split the permuted data into two new samples $X^\prime$ and $\tilde{X}^\prime$ having the same size as the original samples. Then under the null hypothesis, we should have that $MMD(X^\prime,\tilde{X}^\prime)$ be approximately equal to $MMD(X,\tilde{X})$. The corresponding $p$-value can then be calculated by counting how these two quantities are not equal.

### 2022-11-03 paper/sections/related.rmd
incorporated some revsions

A framework for Counterfactual Explanations was first proposed in 2017 by @wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then. Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$. Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $C=\text{cost}(x^\prime)$ incurred by individual $x$ when moving to a counterfactual state $x^\prime$ such that the predicted outcome $M(x^\prime)$ corresponds to some target outcome $y^*$:

>>>>

A framework for Counterfactual Explanations was first proposed in 2017 by Wachter et al. @wachter2017counterfactual and has served as the baseline for most methodologies that have been proposed since then. Let $M: \mathcal{X} \mapsto \mathcal{Y}$ denote some pre-trained model that maps from inputs $X \in \mathcal{X}$ to outputs $Y \in \mathcal{Y}$. Then we are interested in minimizing the cost^[Equivalently, others have referred to this quantity as *complexity* or simply *distance*] $C=\text{cost}(x^\prime)$ incurred by individual $x$ when moving to a counterfactual state $x^\prime$ such that the predicted outcome $M(x^\prime)$ corresponds to some target outcome $y^*$:
